---
title: "R Introduction - Day 2"
output: html_notebook
---

------------------------------------------------------------------------

# IN THIS TOPIC

------------------------------------------------------------------------

-   Built-in data-set
-   Reading Data from Files - CSV, EXCEL, JSON
-   Data Visualization
-   Using Colors
-   Using Packages
-   Projects/Codes Clean up
-   Handling Complex Commands
-   R Markdown
-   Statistics Features: Mean, Median, Regression
-   Further Learning Resources

------------------------------------------------------------------------

# SAMPLE DATA

------------------------------------------------------------------------

load from built-in library (a.k.a base R)

`library("datasets")` or `require("datasets")`

`library()` vs. `require()`

The `library()` by default returns an error if the requested package does not exist. The `require()` is designed to be used inside functions as it gives a warning message and returns a logical value say, FALSE if the requested package is not found and TRUE if the package is loaded.

```{r}
library(help = 'datasets') # display help information about built-in data-sets
?iris # provide help information
str(iris) # show structure information
summary(iris)
dimnames(iris) # returns column names
typeof(iris) # check the storage type
is.data.frame(iris) # check if it's data-frame
class(iris)
head(iris)
(species <- iris$Species) # extracting column/columns
is.data.frame(species)
is.vector(species)
class(species)
levels(species)


```

**basic statistics**

```{r}
max(iris$Sepal.Length)
min(iris$Sepal.Length)
mean(iris$Sepal.Length)
round(mean(iris$Sepal.Length), digits=2)
max(iris$Sepal.Length)
sd(iris$Sepal.Length) # standard deviation

```

------------------------------------------------------------------------

# READING DATA FILES

------------------------------------------------------------------------

To show current working directory, use `getwd()`

To set working directory, use `setwd()` function. Pass a correct absolute path to `setwd()` to change working directory

```{r}
getwd() #  get working directory
```

## reading CSV File

```{r}
allowance <- read.csv("data/allowance.csv") # use relative path here
is.data.frame(allowance)
str(allowance)
summary(allowance)
head(allowance)
ncol(allowance)
nrow(allowance)
dimnames(allowance)
print(allowance) 

```

**reverse the order of rows/columns**

```{r}
nrow(allowance) # returns number of rows
ncol(allowance) # returns number of columns
order(nrow(allowance):1)
(allowance.rows.reversed=allowance[order(nrow(allowance):1), ])
(allowance.cols.reversed=allowance[, order(ncol(allowance):1)])

```

you can also reach out to the environment tab, and click variable name to show the data-frame and GUI manner

**subset() function - let you perform SQL like filtering query**

```{r}
subset(allowance, Basic>120000)

```

**use select parameter to choose columns to retrieve**

```{r}
subset(allowance, Basic>120000, select = c(Assessment_Year, Basic))

```

**use of aggregation functions: e.g.: `min()`, `max()`** When needed, put statements in to multiple lines for readability

```{r}
subset(allowance, 
       Basic==min(Basic), 
       select = c(Assessment_Year, Basic)
       )
subset(allowance, 
       Basic==max(Basic), 
       select = c(Assessment_Year, Basic)
       )

```

## reading Excel File

Reading Excel could be troublesome as there are many version for excel Files. Excel are seldom pure tabular data. Usually embedded with formatting data. To avoid errors of reading excel, you can save/convert it to CSV first

```{r}
installed.packages("tidyverse") # for tidying up data and more function
library(readxl) # requires tidyverse installed first
intakes <- read_excel("data/Students.xlsx", sheet= 1)
summary(intakes)
graduates <- read_excel("data/Students.xlsx", sheet= 2)
str(graduates) # returns the structure of graduates data frame
colnames(graduates) # returns the columns' names
head(graduates)
# alternative approach to retrieve certain columns
data.frame(graduates$AcademicYear, graduates$Undergraduate)
subset(graduates, 
       Undergraduate<10000, 
       select = c(AcademicYear, Undergraduate)
       )
subset(graduates, 
       Undergraduate<10000 & Undergraduate>5000, # multiple conditions 
       select = c(AcademicYear, Undergraduate)
       )
substring(graduates$AcademicYear, 1, 4)
typeof(substring(graduates$AcademicYear, 1, 4))
substring(graduates$AcademicYear, 1, 4) > 2010
as.numeric(substring(graduates$AcademicYear, 1, 4)) - 2000
```

## reading JSON File

JSON file stores data as text in human-readable format. However, JSON files are targeted to read by programs. JSON stands for JavaScript Object Notation

R can read JSON files using the rjson package or others. JSON is very popular to present web api data. \* Example: fixer.io for foreign exchange \* Example: Google Geocoding: for location searching

**install required package**

```{r}
install.packages("rjson")

```

**load required library**

```{r}
library("rjson")
?fromJSON
interbank <- fromJSON(file = "data/daily-figures-interbank-liquidity.json")
interbank
class(interbank)
names(interbank) # return the names of child properties
```

Open up your JSON file in a browser to view JSON data structure. Firefox supports JSON beautifying by default to make it more readable. For Google Chrome, you will have to install extra extension for JSON readability.

## accessing JSON data

```{r}
result <- interbank$result # use $ to access child proproperty
names(result) # return the names of child properties
records = interbank$result$records # json data is nested/multi-level
records # shows the contents
str(records)
class(records)
print(records)

```

It will be a lot easier to read JSON by using add-on. We will see at section of popular packages.

------------------------------------------------------------------------

# DATA VISUALIZATION

------------------------------------------------------------------------

## histograms

```{r}
allowance = read.csv("data/allowance.csv") # use relative path
allowance
hist(allowance$Basic)
# pass extra parameter to control chart's display
hist(allowance$Basic, 
     main = "Basic Allowance",
     xlab = "Basic Allowance for Single Person",
     col="#CCCC00", # RGB color code or color name: red
     border="blue"
     )
```

## line graphs

```{r}
intakes <- read_excel("data/Students.xlsx", sheet= 1)
summary(intakes)
graduates <- read_excel("data/Students.xlsx", sheet= 2)
str(graduates)
head(graduates)
plot(graduates$Undergraduate)
plot(graduates$Undergraduate, 
     type="o"
     )
str(graduates)
plot(graduates$Undergraduate, 
     type="o",
     col = "purple",
     main = "Undergraduate",
     xlab = "Year",
     ylab = "# of student"
     
)

```

Opening helps for `plot` function

```{r}
?plot

```

## bar plot

```{r}
barplot(graduates$Undergraduate)
barplot(graduates$Undergraduate,
        names.arg = graduates$AcademicYear
        )

```

Opening helps for `barplot` function

```{r}
?barplot

```

## pie chart

the following example generates a pie chart of the latest year graduates

```{r}
graduates[1] # returns the first column
typeof(graduates[1])
graduates[1,] # returns the first row
typeof(graduates[1,])
graduates[nrow(graduates),] # returns the last row
graduates[,1] # returns the first row as vector
typeof(graduates[,1]) 
last.row = graduates[nrow(graduates),]
typeof(last.row)
last.row[2:ncol(graduates)] #  excluding the first column - year column
year.18.19 = as.numeric(
  last.row[2:ncol(graduates)]
    ) # excluding the first column - the year column
pie(year.18.19)
colnames(graduates) # all column names
labels = colnames(graduates)[2:ncol(graduates)] # excluding the first column name
pie(year.18.19, labels)

```

## scatter plots

```{r}
library("datasets")
df.iris <- iris
head(df.iris)
plot(x=df.iris$Sepal.Width, 
     y=df.iris$Petal.Width,
     col = "purple"
     )

# specifying a output file
png(file = "output/scatterplot3.png") 

plot(x = df.iris$Sepal.Length,
     y = df.iris$Petal.Length,
     xlab = "Sepal Length",
     ylab = "Petal Length",
     main = "Petal Length vs. Sepal Length",
     col = "purple"
)
dev.off()

```

------------------------------------------------------------------------

# USING COLORS

------------------------------------------------------------------------

```{r}
?colors # show helps on colors 
colors() # display a long list of pre-defined color names

```

## color syntax

```{r}
barplot(graduates$Undergraduate, col = "tomato") # using color name string
barplot(graduates$Undergraduate, col = rgb(.8, .8, 0)) # use rgb() function
barplot(graduates$Undergraduate, col = rgb(200, 0, 0, max=255))
barplot(graduates$Undergraduate, col = "#009900") # use RGB Hex Code String
colors()[400]
barplot(graduates$Undergraduate, col = colors()[400]) # use color index

```

**use TWO colors alternatively**

```{r}
barplot(graduates$Undergraduate, col = c("navy", "orange")) 

```

## using palette

```{r}
?palette
palette()
barplot(graduates$Undergraduate, col = 1:7) # using color #1 to #7 from color palettes

```

**using other built-in palette**

```{r}
barplot(graduates$Undergraduate, col = heat.colors(12)) 

```

## colors for R

[Color for R](https://datalab.cc/rcolors)

## more on using colors for R

[More on using colors for R](https://www.r-bloggers.com/2012/10/palettes-in-r/)

**color tips**

Color is very a dominant design tool to guide user's attention Over-use or mistakenly use of colors becomes confusing. Be careful! Some colors are culture bounded. Red is good for Chinese or east Asia Red is usually NOT good in financial reporting

------------------------------------------------------------------------

# WHAT ARE PACKAGES?

------------------------------------------------------------------------

R provides many free packages (16xxx packages) to add data processing power to your R. R packages are a collection of R functions, complied code and sample data. They are stored under a directory called "library" in the R environment. By default, R installs a set of packages during installation. More packages are added later, when they are needed for some specific purpose. When we start the R console, only the default packages are available by default. Other packages which are already installed have to be loaded explicitly to be used by the R program that is going to use them. Notes: some packages installation might require re-start R studio to take effective.

## packages listing

All the packages available in R language are listed [R Packages](https://cran.r-project.org/web/packages/index.html)

**print the directory where your R packages are stored**

```{r}
.libPaths()

```

## to install a package

syntax:

`install.packages("Package Name")`

## to install multiple packages

syntaxt:

`install.packages("Package1", "Package2")`

**example:** `install.packages("xlsx")`

It will take a while to download requested packages

## to remove a package:

`remove.packages("xlsx")`

------------------------------------------------------------------------

# PACKAGE MANAGER: pacman

------------------------------------------------------------------------

pacman: package manager tool. It makes your package management easier.

```{r}

if (!require("pacman")) install.packages("pacman")
pacman::p_load(pacman, party, psych, rio, tidyverse)

```

**alternative solution for reading csv**

```{r}
(df <- read_csv("data/allowance.csv")) # read_csv is an add-on from tidyverse/readr
?read_csv
?readr
summary(df)
str(df)
head(df)
colnames(df) # returns the column names of a data frame

```

reading another csv: use ing `import()` function

```{r}
df.sal <- import("data/Salaries.csv") # import is a function from add-on
head(df.sal)
colnames(df.sal)

```

## some popular packages

tidyverse, rio, party

**installing tidyvers**

```{r}
installed.packages("tidyverse") # for tidying up data and more function

```

More on tydyverse: [Tidyverse](https://www.tidyverse.org/)

## loading stringr

```{r}
library(stringr)
first.name <- "James"
last.name <- "Bond"
str_glue("My name is {last.name}. {first.name} {last.name}.")

```

**more on string** [stringr](https://stringr.tidyverse.org/)

## importing JSON using add-ons

```{r}
interbank.liquidity <- import("data/daily-figures-interbank-liquidity.json")
summary(interbank.liquidity)
names(interbank.liquidity)
interbank.liquidity$result
```

**another package to load json**

```{r}
library(jsonlite) # load package
interbank = fromJSON("data/daily-figures-interbank-liquidity.json")
class(interbank) # return data.frame
names(interbank)
result <- interbank$result
class(result)
names(result)
records = result$records
class(records) # return data.frame
records
glimpse(records) # Not a built-in function
str(records)
names(records) # show the child properties names
head(records)
subset(records, 
       twi<102, 
       select=c("end_of_date", "hibor_overnight", "hibor_fixing_1m", "twi")
)

```

## access JSON data directly over the web

this will retrieve the live/real-time data

```{r}
url = "https://data.weather.gov.hk/weatherAPI/opendata/weather.php?dataType=fnd&lang=en"
hk.weather <- fromJSON(url)
str(hk.weather)
summary(hk.weather)
hk.weather$generalSituation
hk.weather$updateTime
hk.weather$weatherForecast # returns an array
names(hk.weather) # returns the child properties name
hk.weather$weatherForecast # return all the rows of 9 days forecast
class(hk.weather$weatherForecast) # it's a data frame
hk.weather$weatherForecast[1] # gets a column by column number
hk.weather$weatherForecast$forecastDate # gets a column by column name
hk.weather$weatherForecast[1,] # gets a row by row number
hk.weather$weatherForecast[2,] # gets a row by row number
hk.weather$weatherForecast[1,]$forecastDate # gets a row's property by name
hk.weather$weatherForecast$forecastMaxtemp # returns a columns
hk.weather$weatherForecast$forecastMaxtemp$value
hk.weather$weatherForecast$forecastMaxtemp$value[1]
sprintf("The max tempreature for %s is %s%s.",
        hk.weather$weatherForecast$forecastDate[1], 
        hk.weather$weatherForecast$forecastMaxtemp$value[1], 
        hk.weather$weatherForecast$forecastMaxtemp$unit[1]
        )
```

**Generating Charts**

```{r}
class(hk.weather$weatherForecast)
typeof(hk.weather$weatherForecast)
str(hk.weather$weatherForecast)
barplot(hk.weather$weatherForecast$forecastMaxtemp$value,
     names.arg = hk.weather$weatherForecast$forecastDate,
     col="orange",
     main = "Weather Forecast",
     xlab = "9-Day Forecast",
     ylab = "Degree"
)

```

Exercise: Retrieving live JSON data from HKMA [Daily Interbank Liquidity](https://apidocs.hkma.gov.hk/documentation/market-data-and-statistics/daily-monetary-statistics/daily-figures-interbank-liquidity/)

## exercise: practicing web json

accessing interbank liquidity data from open data

```{r}
hkma.interbank.url = "https://api.hkma.gov.hk/public/market-data-and-statistics/daily-monetary-statistics/daily-figures-interbank-liquidity"
interbank.liquidity = fromJSON(hkma.interbank.url)
# the above retrieval will take a while.  The server response is slow compared to local data file.
summary(interbank.liquidity)
str(interbank.liquidity)
interbank.liquidity$result
str(interbank.liquidity$result)
interbank.records = interbank.liquidity$result$records
head(interbank.records)
interbank.records[1] # returns the first column
interbank.records[1,] # return the first row
interbank.records[1,]$end_of_date # JSON data is nested
interbank.records[1,]$cu_weakside
interbank.records[1,]$cu_strongside

```

**Plotting**

```{r}
plot(interbank.liquidity$result$records$hibor_overnight,
     type="l",
     col="orange",
     main="hibor overnight",
     xlab = "Date",
     ylab = "Rate"
     )

```

**More Finance Data:** [HKMA Open Data](https://data.gov.hk/tc-datasets/provider/hk-hkma/category/finance?order=name&file-content=no)

**More on HK Gov Open Data:** [HK Gov Open Data](https://data.gov.hk/en/)

------------------------------------------------------------------------

# CLEAN UP

------------------------------------------------------------------------

Packages consumes memory. When you are done using some packages, unload them to free memory.

## clean up variables

```{r}
age <- 22
city <- "HK"
ls() # shows the variables in project environment
rm(list = c("age", "city")) # removes "age", "city" variable from memory
rm(list = ls()) # removes all the variables

```

## cleans up loaded packages

```{r}
search() # shows the loaded memory
p_unload(all) # unloads ALL add-ons. Base library will stay
search()
detach("package:datasets", unload=TRUE) # unloads base library

```

## clean console screen

-   press "Clean console" icon
-   keyboard shortcut: CTRL+L
-   or reaching out to the clean button on console window (upper right)

```{r}
cat("\014")

```

------------------------------------------------------------------------

# HANDLING COMPLEX CODES

------------------------------------------------------------------------

## put codes in multiple lines

To improve readability, use together with comment to enable/disable certain parameters easily

```{r}
library("datasets")
df.iris <- iris
plot(x = df.iris$Sepal.Length,
     y = df.iris$Petal.Length,
     # xlab = "Sepal Length", # x label
     # ylab = "Petal Length", # y label
     # main = "Petal Length vs. Sepal Length", # main title
     # col = palette.colors() # color combination
)

```

## piping commands

Nested functions calls becomes really messy and difficult to read especially when each function has multiple parameters.

```{r}
text = "1234.56789" # sometimes number data appears as text when import
round(sqrt(as.double(text)), 2) # can easily miss closing parentheses

```

We need extra packages to enable piping command. Built-in command to load addon: `library("tidyverse")`

We've downloaded pacman (package manager) to make things easier

```{r}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(pacman, rio, tidyverse)

```

Make sure you've load the required addons to run following command

```{r}
as.double(text) %>% sqrt() %>% round(2) 

```

**compare to:** `round(sqrt(as.double(text)), 2)`

**write your long statements in multiple lines to make it more readable**

```{r}
as.double(text) %>% 
  sqrt() %>% 
  round(2) 

```

**unload ALL add-ons**

```{r}
p_unload(all) 

```

------------------------------------------------------------------------

# STATISTICS FEATURES

------------------------------------------------------------------------

```{r}
library("readxl")
intakes <- read_excel("data/Students.xlsx", sheet= 1)
summary(intakes)

```

## general statistic

```{r}
max(intakes$Undergraduate)
min(intakes$Undergraduate)
mean(intakes$Undergraduate)
median(intakes$Undergraduate)

```

## linear regression

Regression analysis is a very widely used statistical tool to establish a relationship model between two variables. One of these variable is called predictor variable whose value is gathered through experiments.

The other variable is called response variable whose value is derived from the predictor variable. Mathematically a linear relationship represents a straight line when plotted as a graph. A non-linear relationship is when exponent of any variable is not equal to 1, which creates a curve.

`y = ax + b` - y is the response variable. - x is the predictor variable. - a and b are constants which are called the coefficients.

**steps for linear regression** 1. Carry out the experiment of gathering a sample of observed values of height and corresponding weight. 2. Create a relationship model using the lm() functions in R. 3. Find the coefficients from the model created and create the mathematical equation using these 4. Get a summary of the relationship model to know the average error in prediction. Also called residuals. 5. To predict the weight of new persons, use the predict() function in R.

## example: predicting the weight

```{r}
x <- c(151, 174, 138, 186, 128, 136, 179, 163, 152, 131) # height
y <- c(63, 81, 56, 91, 47, 57, 76, 72, 62, 48) # weight

```

`lm()` function This function creates the relationship model between the predictor and the response variable

```{r}
relation <- lm(y~x) # or lm(formula = y ~ x)
print(relation)
summary(relation)
typeof(relation)
class(relation)

```

Find weight of a person with height 175 Use predict() fucntion for predicting

```{r}
h <- data.frame(x=175)
h
(w <- predict(relation, h))

h2 <- data.frame(x=c(175, 180, 190))
h2
(w2 <- predict(relation, h2))
```

**plot the regression**

```{r}
plot(y,x,
     col = "#009900",
     main = "Height & Weight Regression",
     abline(lm(x~y)),
     cex = 1.3,
     pch = 16,
     xlab = "Weight in Kg",
     ylab = "Height in cm"
     )
```

**More examples on R Statistics** at [Tutorials Point](https://www.tutorialspoint.com/r/index.htm)

## Further reference

-   [tydyverse](https://www.tidyverse.org/)

-   [tydyverse e-book online](https://r4ds.had.co.nz/)

-   [R Packages](https://cran.r-project.org/web/packages/index.html)
