---
title: "R Introduction - Day 1"
output: html_notebook
---

------------------------------------------------------------------------

# INTERACTING WITH R-NOTEBOOK

------------------------------------------------------------------------

This is a [R Markdown](http://rmarkdown.rstudio.com) Notebook. When you execute code within the notebook, the results appear beneath the code.

## To run a chunk of codes

Try executing this chunk by clicking the **Run** button (the GREEN play button) within the chunk or by placing your cursor inside it and pressing **CTRL+SHIFT+ENTER** (for Windows) or **CMD+SHIFT+ENTER** (for Mac).

```{r}
plot(cars)
```

## To run codes line by line

Press **CTRL+ENTER** (Windows) or **CMD+ENTER** (Mac) to run the current line and the cursor will move alone to the next line in the chunk.

```{r}
print(100)
print(200)
print(300)
print(400)
print(500)
```

## To insert a new chunk of codes

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing **CTRL+Alt+I** (Windows) or **CMD+OPTION+I** (Mac).

## To preview a notebook output

When you save the notebook, an HTML file containing the codes and outputs will be saved alongside it (click the **Preview** button or press **CTRL+SHIFT+K** or **CMD+SHIFT+K** to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike **Knit**, **Preview** does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

------------------------------------------------------------------------

# TODAY'S TOPICS

------------------------------------------------------------------------

-   Hands on R
-   Basic Syntax
-   Using Help
-   Variables
-   Data Types
-   Using Comments
-   Operators
-   Strings
-   Vectors, Lists, Matrices
-   Data Frames
-   Condition Control
-   Looping
-   Functions
-   Built-in Sample Data Sets

------------------------------------------------------------------------

# A FEW USEFUL SHORTCUTS

------------------------------------------------------------------------

Copy Line(s): **SHIFT+ALT+UP/DOWN** (Windows) or **CMD+OPTION+UP/DOWN** (Mac)

Move Line(s) Up/Down: **ALT+UP/DOWN** to move line(s) up or down

More useful keyboard shortcuts: [R Shortcuts](https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts)

------------------------------------------------------------------------

# GETTING HELPS

------------------------------------------------------------------------

To show helps, adding a question mark `?` in front of a command

```{r}
?toupper
?is.vector
```

------------------------------------------------------------------------

# ADDING COMMENTS TO YOUR CODES

------------------------------------------------------------------------

Use comment symbol `#` to add simple explanation to your codes or to temporarily disable certain lines of codes

shortcut for commenting/un-commenting

On Windows: **CTRL+SHIFT+C**

On Mac: **COMMAND+SHIFT+C**

```{r}
# this is comment. the lines below print numbers of various value
print(100) # this is also comment. this line print 100
print(200) 
# the following line is comment out and therefore it is NOT executed.
# print(300) 
print(400)
print(500)
```

------------------------------------------------------------------------

# DECLARING VARIABLES

------------------------------------------------------------------------

Variables are nothing but reserved computer memory locations to store values for later computation. When you create a variable, you reserve some space in memory to store value. A valid variable name consists of **letters**, **numbers**, **dot** or **underline**

-   A variable name starts with a letter or dot
-   A variable name **CANNOT** start with `_` or number
-   It's case sensitive

Valid variable names: `first_name`, `first.name`, `firstname`, `address2`, `.address`

Invalid variable names: `_mobile`, `2mobile`, `.2mobile`

```{r}
my.name <- "April"
MY.NAME <- "Benny"
my_name <- "Cathy"
```

------------------------------------------------------------------------

# VARIABLES ASSIGNMENT

------------------------------------------------------------------------

R allows 5 forms of assignment operators

**leftward assignment**: value on the right are stored to variable name on the left.

```         
x <- value
x = value
x <<- value
```

**rightward assignment**: value on the left are stored to variable name on the right.

```         
value -> x
value ->> x
```

```{r}
city1 <- "Beijing"
city2 = "Shanghai"
city3 <<- "Guangzhou"
"Shenzhen" -> city4
"Qingdao" ->> city5
(city6 <- "Hangzhou") # add a pair of parentheses to declare a variable and print it out to the screen
```

## list variables in environment

```{r}
print(ls()) # list all variables
print(ls(pattern="city")) #list variables with name start 'city'
```

## clear all variables

```{r}
rm(city6)
# rm(list =ls()) # remove all variables
```

------------------------------------------------------------------------

# PRINTING OUT

------------------------------------------------------------------------

## use `print()` function

```{r}
print(city1)
cat(city1, city2, city3, city4, city5)
```

## use `sprintf()` function for c style printing

more about [sprintf](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/sprintf)

```{r}
first.name <- "James"
(last.name <- "Bond") # define a variable and print it's value to console
code <- 7L # L is added at end to indicate is a whole number
sprintf("My namne is %s. %s %s. I am 00%i.",
        last.name, first.name, last.name, code)
# like above, you can put a statement to multiple lines
```

------------------------------------------------------------------------

# DATA TYPES

------------------------------------------------------------------------

*Vectors: Logical, Numeric, Integer, Complex, Character, Raw* Lists \* Matrices \* Arrays \* Factors \* Data Frames

Types in R are rather different to other programming language.\
It could be confusing at first.

```{r}
is.student <- TRUE # logical
height <- 182.5 # double
age <- 20L # integer - add an ending 'L' to differentiate to double
c <- "A" # character
book.title <- "Forrest Gump" # character
bytes <- charToRaw("ABC") # raw bytes of ASCII codes

```

vector: a series of values

an item in vector is called a component

use `c()` function to combine multiple components

```{r}
fruits <- c("apple", "banana", "orange")
```

`is.vector()` checks if the tested object (obj) is a vector or expression that has either no attributes or only the names attribute set

```{r}
is.vector(is.student) # returns TRUE
is.vector(age) # returns TRUE
is.vector(fruits) # returns TRUE
is.atomic(fruits) # returns TRUE
is.list(fruits) # returns FALSE
```

## upper-case / lower-case converstion

About [ASCII Codes](https://en.m.wikipedia.org/wiki/File:ASCII-Table-wide.svg)

```{r}
charToRaw("A") # outputs 41 hex value
charToRaw("a") # outputs 61 hex value
as.numeric(charToRaw("A")) # outputs 65 decimal value
as.numeric(charToRaw("a")) # outputs 97 decimal value
```

The value difference for lower case to upper case

97 - 65 = 32

```{r}
bytes <- charToRaw("ABC") # raw bytes of ASCII codes: 41 42 43
print(bytes) # print hexidecimal value: 41 42 43
print(as.numeric(bytes)) # print decimal value: 65 66 67
as.numeric(bytes)+32 # outputs : 97 98 99 (decimal values)
bytes_converted = as.raw(as.numeric(bytes)+32) 
bytes_converted # hex value 61 62 63
rawToChar(bytes_converted) # outputs lower case: abc
```

## Exercise

convert your name in lower case to upper case

Hints: refer to the codes logic above

```{r}
my.name = "sunny"

```

------------------------------------------------------------------------

# CHECKING TYPES: `typeof()`, `class()`, `mode()`

------------------------------------------------------------------------

`typeof()` determines the (R internal) type or storage mode of any object

`class()` returns the values of the class attribute of an R object

`mode()` the mode specifies how the object is stored in memory, same to typeof()

```{r}
typeof(is.student)
typeof(bytes)
class(is.student)
class(bytes)
class(fruits)
typeof(fruits)
mode(fruits)
is.vector(fruits) # check if it is a vector
class(typeof) # this returns 'function'
typeof(typeof) # this returns 'closure'
```

------------------------------------------------------------------------

# LIST

------------------------------------------------------------------------

A list is an R-object which can contain many different types of elements. it could include vectors, functions and even another list inside it.

```{r}
my.list <- list(fruits, height, age)
names(my.list) <- c("fruites", "height", "age")
my.list
is.atomic(my.list) # returns FALSE
is.list(my.list) # returns TRUE
my.list[1] # use bracket (aka square bracket) to specify position of element
my.list[2] # index starts at 1
my.list[4] # index out of range, returns NULL, means nothing

length(my.list) # returns 3
typeof(my.list) # returns 'list'
typeof(my.list[1]) # returns 'list' a sub-item return in the form of list 
typeof(my.list[[1]]) # use double bracket to get the contents. returns 'character'

my.list$fruites[1] # gets an item as sub-list
my.list$fruites[[1]] # gets an item as content
my.list$fruites
my.list$fruites[1]

```

------------------------------------------------------------------------

# RANGE

------------------------------------------------------------------------

```{r}
r = 1:10
r
r[1]
length(r)
r[length(r)]
(r2 = seq(1, 10, by=2))
(r3 = r2 * 10) # each element in r2 times 10 and stored in a new range
```

------------------------------------------------------------------------

# MATRICES

------------------------------------------------------------------------

A matrix is a two-dimensional rectangular data set. It can be created using a vector input to the matrix function.

```{r}
m1 = matrix( c(1:12), nrow=3) # nrow is number of row
m1
m2 = matrix( c(1:12), nrow=3, byrow = TRUE) # default processing is by col
m2
m3 = matrix( c(1:12), ncol=2)
m3
```

------------------------------------------------------------------------

# ARRAYS

------------------------------------------------------------------------

Matrices are confined to two dimensions. Arrays can be of any number of dimensions. The array function takes a `dim` attribute which creates the required number of dimension.

In the following example, we create an array with two elements and each is 3x3 matrices.

```{r}
myarray <- array(c(1:18), dim = c(3, 3, 2))
myarray
myarray[,,1] # use comas to separate position number of various dimension
myarray[,,2]
myarray[1,1,2]
```

------------------------------------------------------------------------

# FACTORS

------------------------------------------------------------------------

Factors are the r-objects which are created using a vector.

Factors are created using the `factor()` function.

the `nlevels()` functions gives the count of levels (number of distinct values)

```{r}
cities <- c('Shenzhen','Beijing','Shanghai','Shenzhen','Qingdao','Guangzhou','Beijing')
factor_cities <- factor(cities)
factor_cities
levels(factor_cities)
nlevels(factor_cities) # returns 5
```

------------------------------------------------------------------------

# DATA FRAME

------------------------------------------------------------------------

-   data frames are tabular data objects.
-   each column can contain different type of data.
-   the column names should be non-empty.
-   the row names should be unique.
-   the data stored in a data frame can be of numeric, factor or character type.
-   each column should contain same number of data items.

Data Frames are created using the `data.frame()` function. We don't normally create data frame like the following example. Instead, we usually import data from external file such as **csv**, **json**, **excel**, **xml**

```{r}
name <- c("April", "Ben", "Catherine", "Debra", "Eric", "Frankie", "George")
age <- c(20, 21, 19, 22, 23, 24, 25)
is.student <- c(TRUE, FALSE, T, F, F, F, F)
df = data.frame(name, age, is.student)
df
head(df) # list the first SIX rows
colnames(df) # show the column names of data-frame
```

## information about data-frame

```{r}
str(df) # prints the structure of a data-frame
summary(df) # prints the summary information of a data-frame
nrow(df) # returns number of row
ncol(df) # returns number of column
length(df) # number of variables, same as number of columns

```

## type information of data frame

```{r}
class(df) # returns data.frame
typeof(df) # return list
mode(df) # return list
```

## ise `$` symbol to refer to column

```{r}
df$name # use '$' sign to indicate column name getting the 'name' column
df$age # getting the age column
```

## getting row/column

```{r}
df[1,] # get the first row
is.vector(df[1,])
class(df[1,])
df[,1] # get the first col
df[1] # get the first col, return as data.frame
```

The above seems to give identical result, their types are different. Let's check the return types.

```{r}
cat(class(df[,1]), class(df[1]))
cat(typeof(df[,1]), typeof(df[1]))
cat(is.vector(df[,1]), is.vector(df[1]))
cat(is.data.frame(df[,1]), is.data.frame(df[1]))
df[1,1] # get the first col in the first row
```

## extract columns from data frame

```{r}
data.frame(df$name, df$is.student) # take two column and form a new data frame

```

## extra first two rows

```{r}
df[1:2,]

```

## extra first two rows, first two columns

use `:` to define a range with start number and ending number

```{r}
df[1:2, 1:2]

```

------------------------------------------------------------------------

# OPERATORS

------------------------------------------------------------------------

-   Arithmetic Operators
-   Relational Operators
-   Logical Operators
-   Assignment Operators

## arithmetic operators

```{r}
v1 = c(1, 2, 3, 4)
v2 = c(100, 200, 300, 400)
print(v1+10) # add 10 to each element
print(v1-4)
print(v1*10)
print(v1/2)
print(v1+v2) # two vectors must be same length
print(v1-v2)
print(v1*v2)
print(v1/v2)
print(v2%%(v1+9))
print(v2^v1)

```

## relational operators

Each element of the first vector is compared to the corresponding element of the second vector. The result of each comparison is a Boolean value.

```{r}
v1 = c(1,   3,   5, 7, 9)
v2 = c(100, 300, 5, 1, 1)
print(v1>v2) # one to one element comparison for two list
print(v1<v2)
print(v1==v2)
print(v1!=v2)
print(v1>=v2)
print(v1<=v2)
print(v1>7) # compare each element from the list to simple value 7

```

## logical operators

-   `&` : and
-   `|` : or
-   `!` : not

```{r}
v3 = c(T, T, T, F, F)
v4 = c(F, F, T, F, F)
print(v3&v4)
print(v3|v4)
print(!v3)
print(!v4)

```

**when the length of two vector is not the same, the shorter vector will replicate itself to match the length**

```{r}
v5 <- c(100, 200, 300, 400, 500, 600)
v6 <- c(1,2)
print(v5+v6) # v6 becomes c(1,2,1,2,1,2) while performing the operation

```

**filtering with logical operator**

```{r}
r = 1:100
r
(r>50) # generates a list of TRUE or FALSE values
(r%%2)
(r%%2==0)
r[r%%2==0] # returns all the elements that are dividable by 2
r[r>50] # returns all the elements that are greater than 50
r[r%%2==0 & r>50] # multiple conditions
```

## assignment operators

-   leftward assignment: `=`, `<-`, `<<-`
-   rightward assignment: `->`, `->>`

**sequence operator** `:`

```{r}
s = 1:10
s

```

**check if an element exists: use `%in%`**

We previously defined: `fruits <- c("apple", "banana", "orange")`

```{r}
"apple" %in% fruits # returns TRUE
"pear" %in% fruits # returns FALSE

```

------------------------------------------------------------------------

# PROGRAM CONTROL: CONDITION

------------------------------------------------------------------------

```{r}
is.student = F
if (is.student) {
  print("20% OFF")
} else {
    print("No. Discount.")
}

```

**you can further chain up "else if" for more complex conditions checking**

```{r}
number <- 101
if(number >= 1000) {
  print("it's a 4 digit number or above")
} else if (number>=100) {
  print("it's a 3 digit number ")
} else if (number>10) {
  print("it's a 2 digit number ")
} else {
  print("it's a 1 digit number ")
}

```

------------------------------------------------------------------------

# PROGRAM CONTROL: LOOPING

------------------------------------------------------------------------

## for loop

```{r}
for (fruit in fruits){
  print(fruit)
}
```

## while loop

```{r}
number <- 0
while (number < 10) {
  print(number)
  number = number + 1
}
```

------------------------------------------------------------------------

# FUNCTION

------------------------------------------------------------------------

## built-in functions

```{r}
r = 1:100
print(max(r)) # maximum value
print(min(r)) # minimal value
print(mean(r)) # mean value
print(sum(r)) # summation

```

## string function `paste()`

```{r}
a <- "How"
b <- 'Are'
c <- "You"
print(paste(a,b,c))
print(paste(a,b,c, sep = "-"))

```

## convert text to upper-case or to lower-case

```{r}
my.name <- "Johnny"
toupper(my.name) # to upper case
tolower(my.name) # to lower case
nchar(my.name) # number of character in the string
substring(my.name, 1 , 3) # returns "Joh"
substring(my.name, 4) # returns "nny"

```

## formatting numbers

```{r}
salary = 12345.6789
typeof(salary)
format(salary, digits=5) # gives 12346
format(salary, digits=6) # gives 12345.7
format(salary, nsmall=2) # gives 12345.68
?format

```

## `c()` function / combine function

```{r}
about.me = c("Johnny", 25, TRUE)
about.me # number and boolean are converted to characters
about.me[1] # index starts at 1

```

## define your own function

```{r}
mysquare <- function(number) {
  return (number * number)
}
print(mysquare(3))

```

## filtering vectors

```{r}
days <- c("Sun","Mon","Tue","Wed","Thurs","Fri","Sat")
(selected_days = days[c(1,3,5)])
(selected_days = days[c(T,F,F,F,F,F,T)])
(selected_days = days[c(-1,-7)]) # use '-" sign to indicate exclusion of element

```

## sorting vector

```{r}
sort(days) # alphabet sorting
sort(days, decreasing = TRUE) # alphabet sorting
numbers = c(1,9,5,7,3)
sort(numbers)
sort(numbers, decreasing = TRUE)

```

------------------------------------------------------------------------

# SAMPLE DATASET

------------------------------------------------------------------------

```{r}
library(datasets) # load dataset from built-in library
iris
typeof(iris)
?iris
summary(iris)
head(iris)
hist(iris$Sepal.Width)
plot(iris)

```

------------------------------------------------------------------------

# PREPARE FOR NEXT SESSIONS

------------------------------------------------------------------------

-   We will talk about packages next session
-   Downloading packages will take some times, so run the following command to download packages that we will need for next sessions in advance

```{r}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(pacman, party, psych, rio, tidyverse)

```

You will see console displaying packages download one by one
